---
title: "R & RStudio Setup"
author: "Ozan Ozbeker"
execute: 
  eval: false
format: gfm
---

# Intro

This is a hybrid blog post/README/code file where it is the actual code file I use when I update my R environment/computer, but also I wrote some descriptions along the way so if somebody else stumbles upon this and likes what they see, they can copy it or maybe even find something new that they didn't know about before.

# .Rprofile

The `.Rprofile` file is an R script that runs each time R starts up that lets you to customize your R environment. It can be used to set global options, load packages, define functions, and customize the R startup process. Or in my case, have some fun ;)

## Options

`options()` lets you set the global options for R. All of them have a default, and these are the ones I've changed for my preference. You can use `?options` to view the full list.

```{r filename=".Rprofile"}
#| label: my-options

options(
  scipen = 999999,               # A penalty to be applied to turning regular values to scientific notation
  setWidthOnResize = TRUE,       # If set and TRUE, the terminal resize when the terminal pane is resized
  useFancyQuotes = FALSE,        # Turn off fancy quotes
  warn = 1,                      # Prints warnings as they happen instead of when the top function ends.
  warnPartialMatchArgs = TRUE,   # Warns if partial matching is used in argument matching.
  warnPartialMatchAttr = TRUE,   # Warns if partial matching is used in extracting attributes via attr.
  warnPartialMatchDollar = TRUE, # Warns if partial matching is used for extraction by $.
  prompt = 'R> ',                # This is complete personal preference
  continue = '+> '               # This is complete personal preference
)
```

## R Completion Options

The `rc.settings` function is part of the utils package in R, which configures settings for the R completion mechanism. It can enable or disable specific completion features like inter-package completion, function argument completion, and fuzzy matching. You can use `?rc.settings` to view the full list.

1.  `ipck` (Inter-Package Completion):

    -   Description: This option enables or disables completion of object names across all loaded packages.
    -   Explanation: If you work with multiple packages, you can quickly access functions and objects from any loaded package without having to remember which package they belong to.

2.  `func` (Function Argument Completion):

    -   Description: This option enables or disables completion of function argument names.
    -   Explanation: This helps you to write function calls more efficiently by providing argument suggestions, reducing the need to look up function documentation frequently.

3.  `fuzzy` (Fuzzy Matching):

    -   Description: This option enables or disables fuzzy matching for completion.
    -   Explanation: This is useful when you cannot recall the exact name of a function or object. Fuzzy matching will suggest completions that closely resemble what you have typed, saving time and reducing frustration.

In short, the more help I can get coding, the better ðŸ˜

```{r filename=".Rprofile"}
#| label: my-rc-settings

rc.settings(ipck = TRUE, func = TRUE, fuzzy = TRUE)
```

## Interactive Sessions

You can use this code section if you want things to enable in an interactive session, like auto-loading certain packages. Some people use it to load packages that they always use, like `tidyverse`, `usethis`, `devtools`, and so on, but I'm in the camp that every script should be very explicit in what packages are being used, so I use this feature to have some fun.

If in an interactive session, this code will output my message in the console everytime R starts. I've always wanted my own [J.A.R.V.I.S.](https://en.wikipedia.org/wiki/J.A.R.V.I.S.), R is close enough ðŸ¥²

```{r filename=".Rprofile"}
#| label: my-startup-message

if (interactive()) {
  message("Welcome Mr. Ozbeker, how may I be of assistance?")
}
```

# r.snippets & rstudio_bindings.json

RStudio lets you add custom snippets (pieces of code that can be quickly inserted, useful for repetitive code) and key bindings that can really help your code flow if you use them. For both files, I only added one thing, but they've really improved my coding experience.

```{r filename="r.snippets"}
#| label: my-custom-snippets

snippet user
  user = Sys.info()[['user']]
```

```{json filename="rstudio_bindings.json"}
{
    "insertPipeOperator": "Ctrl+\\"
}
```

# r-rstudio-setup.qmd

Last but not least, we have this Quarto file that applies everything above as well as some custom code we will see below. As I write this, it feels a little 4th wall breaking, but hey that's life ðŸ˜Ž

## Packages

These are the packages I most frequently use, loosely grouped into the categories below. This does not include dependency packages. So for example, I use `quarto` to render Quarto documents via R code, but I don't directly use the `markdown` package myself, so it is not listed below.

You can find the info page for each package at `https://cran.r-project.org/web/packages/[package_name]/index.html` where `[package_name]` is the name of the package.

```{r filename="r-and-rstudio-setup.qmd"}
#| label: install-packages

install.packages(c(
  # Meta Packages
  'tidyverse',  # Easily Install and Load the "Tidyverse" packages | https://tidyverse.tidyverse.org
  'tidymodels', # Easily Install and Load the "Tidymodels" packages | https://tidymodels.tidymodels.org/
  
  # Programming & Development
  'purrr',    # Functional Programming Tools | https://purrr.tidyverse.org/
  'forcats',  # Tools for Working with Categorical Variables (Factors) | https://forcats.tidyverse.org/
  'keyring',  # Package for accessing OS's credential store | https://keyring.r-lib.org/
  'fs',       # Cross-Platform File System Operations Based on 'libuv' | https://fs.r-lib.org/

  # Import
  'readr',    # Read Rectangular Data | https://readr.tidyverse.org/
  'httr2',    # Perform HTTP Requests and Process the Responses | https://httr2.r-lib.org/
  'readxl',   # Read Excel Files | https://readxl.tidyverse.org/
  'rvest',    # Easily Harvest (Scrape) Web Pages | https://rvest.tidyverse.org/
  'jsonlite', # A Simple and Robust JSON Parser and Generator for R | https://jeroen.r-universe.dev/jsonlite
  
  # Tidy/Transform
  'dplyr',     # A Grammar of Data Manipulation | https://dplyr.tidyverse.org/
  'tidyr',     # Tidy Messy Data | https://tidyr.tidyverse.org/
  'tibble',    # Simple Data Frames | https://tibble.tidyverse.org/
  'stringr',   # Simple, Consistent Wrappers for Common String Operations | https://stringr.tidyverse.org/
  'lubridate', # Make Dealing with Dates a Little Easier | https://lubridate.tidyverse.org/
  'janitor',   # Simple Tools for Examining and Cleaning Dirty Data | https://sfirke.github.io/janitor/index.html 

  # Visualize
  'ggplot2', # Create Elegant Data Visualizations Using the Grammar of Graphics | https://ggplot2.tidyverse.org/
  
  # Model
  'rsample',   # General Re-sampling Infrastructure | https://rsample.tidymodels.org/
  'parsnip',   # A Common API to Modeling and Analysis Functions | https://parsnip.tidymodels.org/
  'recipes',   # Pre-processing and Feature Engineering Steps for Modeling | https://recipes.tidymodels.org/
  'workflows', # Modeling Workflows | https://workflows.tidymodels.org/
  'tune',      # Tidy Tuning Tools | https://tune.tidymodels.org/
  'yardstick', # Tidy Characterizations of Model Performance | https://yardstick.tidymodels.org/
  'broom',     # Convert Statistical Objects into Tidy Tibbles | https://broom.tidymodels.org/
  'dials',     # Tools for Creating Tuning Parameter Values | https://dials.tidymodels.org/
  'infer',     # Tidy Statistical Inference | https://infer.tidymodels.org/
  'corrr',     # Correlations in R | https://corrr.tidymodels.org/
  
  # Communicate
  'quarto', # R Interface to 'Quarto' Markdown Publishing System

  # Database
  'DBI',     # R Database Interface | https://dbi.r-dbi.org/index.html
  'odbc',    # Connect to ODBC Compatible Databases (using the {DBI} Interface) | https://odbc.r-dbi.org/
  'dbplyr',  # A {dplyr} Back End for Databases | https://dbplyr.tidyverse.org/
  'duckdb',  # {DBI} Package for the DuckDB Database Management System | https://r.duckdb.org/
  'duckplyr' # A DuckDB-backed Version of {dplyr} | https://duckdblabs.github.io/duckplyr/
))

# Other packages I'm interested in:
# 'profvis'  # Interactive Visualizations for Profiling R Code | https://rstudio.github.io/profvis/
# 'targets'  # Pipelining Tools in R | https://docs.ropensci.org/targets/
# 'testthat' # Unit Testing for R | https://testthat.r-lib.org
# 'usethis'  # Automate Package and Project Setup | https://usethis.r-lib.org/
# 'devtools' # Tools to Make Developing R Packages Easier | https://devtools.r-lib.org/
# 'zoo'      # S3 Infrastructure for Regular and Irregular Time Series (Z's Ordered Observations) | https://zoo.R-Forge.R-project.org/
```

## RStudio Configuration

As I've been using R & RStudio, I've learned which settings I like the most, whether they are RStudio settings or R options, as well as some custom snippets. This section of code takes the respective files and copies them to where the base files for R and RStudioa are. I really created this because of my job where we log into different virtual machines, and I got sick of resetting all of these options by hand, so now I just pull this repo from GitHub and run the code.

The directories don't exist by default, they are created when you manually change the respective settings, so this chunk takes care of that, especially if it's a fresh install of RStudio.

```{r filename="r-and-rstudio-setup.qmd"}
#| label: create-rstudio-directories

purrr::walk(
  .x = c('themes', 'keybindings', 'snippets'),
  .f = \(directory) stringr::str_c("C:/Users/", {Sys.info()[['user']]}, "/AppData/Roaming/RStudio/", directory) |> fs::dir_create()
)
```

Then we copy the files from this folder to their appropriate locations.

```{r filename="r-and-rstudio-setup.qmd"}
#| label: copy-files

purrr::pwalk(
  .l = tibble::tribble(
    ~file,                   ~destination,
    'rstudio_bindings.json',  stringr::str_glue("C:/Users/{Sys.info()[['user']]}/AppData/Roaming/RStudio/keybindings/rstudio_bindings.json"),
    'KISS - OZBEKER.rstheme', stringr::str_glue("C:/Users/{Sys.info()[['user']]}/AppData/Roaming/RStudio/themes/KISS - OZBEKER.rstheme"),
    'r.snippets',             stringr::str_glue("C:/Users/{Sys.info()[['user']]}/AppData/Roaming/RStudio/snippets/r.snippets"),
    'rstudio-prefs.json',     stringr::str_glue("C:/Users/{Sys.info()[['user']]}/AppData/Roaming/RStudio/rstudio-prefs.json"),
    '.Rprofile',              stringr::str_glue("C:/Users/{Sys.info()[['user']]}/Documents/.Rprofile")),
  .f = \(file, destination) fs::file_copy(path = file, new_path = destination, overwrite = TRUE)
)
```

## "Real" Dark Mode & Theme

I hate that when you use dark mode in RStudio, it's just this dark blue shade for the UI, not actually dark/black. Also, I'm really not a fan of any of the themes that come with RStudio. Luckily for me, GitHub user [rileytwo](https://github.com/rileytwo) made a fix for both of those problems for me.

-   [{darkstudio}](https://github.com/rileytwo/darkstudio) is a package that turns the IDE elements to shades of black whenever you are using a theme with `rs-theme-is-dark: TRUE`.
-   [Kiss: Keep It Stupid Simple](https://github.com/rileytwo/kiss) is a theme from Riley that just looks so good, it's been my go to for over a year now.

If either of these interest you, please go check them out as they have been the staple to my R experience and I like to support the creator as best as I can.

```{r filename="r-and-rstudio-setup.qmd"}
#| label: activate-dark-studio

devtools::install_github('rileytwo/darkstudio')
darkstudio::activate() # This requires admin privileges
```
